---
title: "Coursera Practical Machine Learning Course Project"
author: "Bob Currie"
date: "December 18, 2017"
output: html_document
output_dir: "."
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Qualitative Activity Recognition (Prediction) of Weight Lifting Exercises

## Synopsis
The data used for this project was obtained from a study conducted by
see: 
The study used trackers to obtained measurements from 6 diffrent participants as they correctly and incorrectly performed sets of dumbbell curls. The readings from the trackers were collected for a series of rolling time windows of (explain here). The result was 19602 observations with 160 features.

Using a random forest model, we see that .

When tested on a testing set withheld from the data, 


## Data Processing

Download the datasets from the web and then read them into R dataframes

Note: There are two files associated with this project. "pml-training.csv" and "pml-testing.csv". It is important to note that "pml-testing" is used ONLY for the Prediction Quiz portion. The "pml-training" dataset will be read into an R dataframe (**qarwle**), then split into two datafames; qarwle_train (for developing and training the model), and qarwle_test (for measuring out of sample model error and accuracy)

```{r cars}
summary(cars)

## Download files from web
linkTraining <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
linkTesting <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"

# Download to working directory
download.file(url = linkTraining, destfile = "pml-training.csv")
download.file(url = linkTesting, destfile = "pml-testing.csv")
```

### Data Pre-processing and Tidying

### Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.







## Results
